Deriving an Executable Model of a Data Stream Protocol from its Specification

### 2009-08-04 papers/ModelBasedTestCaseGenerationDataStreamProtocol/paper.tex
ModelBasedTestCaseGenerationDataStreamProtocol paper: added packet structure description

%\usepackage{todonotes}

% \usepackage{html}
% \usepackage{url}
\usepackage{xspace}
%\usepackage{doublespace}
\usepackage{tabularx}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage{stmaryrd}
% \ifpdf
% \usepackage[centredisplay]{diagrams}
% \else
% \usepackage[centredisplay,PostScript=dvips]{diagrams}
% \fi
\usepackage{float}

@@

\newcommand{\tablesize}{\footnotesize}
\newcommand{\eg}{e.g.,\xspace}
\newcommand{\ie}{i.e.,\xspace}
\newcommand{\etc}{etc.\xspace}
\newcommand{\myhref}[2]{\ifpdf\href{#1}{#2}\else\htmladdnormallinkfoot{#2}{#1}\fi}
%\newcommand{\myhref}[2]{\emph{#2}}
\newcommand{\todo}{\textbf{TODO:}}
\newcommand{\ST}{\emph{SenseTile}\xspace}
\newcommand{\STs}{\emph{SenseTiles}\xspace}
\newcommand{\STS}{\emph{SenseTile Simulator}\xspace}
\newcommand{\datastore}{\STs Scientific Datastore\xspace}
\newcommand{\computefarm}{The \STs Scientific Compute Farm\xspace}
\newcommand{\computefarmlong}{UCD CASL \STs Software and Data Compute Server Farm\xspace}
\newcommand{\sensorfarm}{The \STs\xspace}

@@

  Sensor packages typically communicate (control sensors and report
  sensor measurements) with a streaming packet-based network protocol.
  The \ST device is an example of such a sensor package, as it
  communicates with an arbitrary computer system (running Linux or
  Windows, at the moment) over a USB bus using a streaming
  packet-based protocol. 
  This sensor platform is a custom-designed,
  but inexpensive, sensor platform (called the \ST) paired with
  general-purpose small-scale compute nodes, including everything from
  low-power PDAs to powerful touchscreen-based portable computers.
  There are over a dozen sensors on the \ST itself, and new sensors
  can be added by adding them to the USB bus.
  A third-party low-level
  library, provided by the USB chip vendor, creates and parses
  generic USB communication packets, while a custom medium-level Java
  driver creates and parses \ST streaming communication packets. 
  This
  latter Java driver is formally specified, validated with rigorous
  unit testing, and verified using extended static checking. 
  Because of this process and combination of technologies, using this
  specification we are able to both communicate with a real \ST
  \textbf{and} generate a formally specified, executable, functional
  model-based specification for the protocol itself written in the
  Java Modeling Language.
  \todo executable? how? what does this mean?
  This protocol model is itself an executable
  stream simulator that can be used to develop new \ST user
  applications without having the \ST hardware at all. 
  Consequently,
  from this single specification, we can test that the \ST hardware
  communicates as expected, test that our \ST simulator communicates
  as expected, and simulate the \ST protocol stream itself.

@@

@@

\todo describe package

>>>>

\usepackage{xspace}
\usepackage{tabularx}
\usepackage{epsfig}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{eucal}
\usepackage{stmaryrd}
\usepackage{float}

@@

\newcommand{\tablesize}{\footnotesize}
\newcommand{\eg}{e.g.,\xspace}
\newcommand{\ie}{i.e.,\xspace}
\newcommand{\etc}{etc.\xspace}
\newcommand{\myhref}[2]{\ifpdf\href{#1}{#2}\else\htmladdnormallinkfoot{#2}{#1}\fi}
\newcommand{\myhreffootnote}[3]{\myhref{#1}{#2}\footnote{#3 \myhref{#1}{#1}}}

\newcommand{\notev}[1]{\xspace$\textcolor{blue}{\omega^\textsf{vieri}}$\marginpar{\scriptsize\textsf{Vieri:} #1}}
\newcommand{\notej}[1]{\xspace$\frac{\varocircle}{\textsf{jk}}$\marginpar{\scriptsize\textsf{Joe:} #1}}
\newcommand{\noted}[1]{\xspace$\textcolor{red}{\dagger^\textsf{dragan}}$\marginpar{\scriptsize\textsf{Dragan:} #1}}

\newcommand{\todo}{\textbf{TODO:}}

\newcommand{\ST}{\emph{SenseTile}\xspace}
\newcommand{\STSB}{\ST \emph{Sensor Board}\xspace}
\newcommand{\STSBs}{\ST \emph{Sensor Boards}\xspace}
\newcommand{\STPU}{\ST \emph{Processor Unit}\xspace}
\newcommand{\STPUs}{\ST \emph{Processor Units}\xspace}
\newcommand{\STU}{\ST \emph{Unit}\xspace}
\newcommand{\STUs}{\ST \emph{Unit}\xspace}

\newcommand{\STs}{\emph{SenseTiles}\xspace}

\newcommand{\simulator}{\STSB \emph{Simulator}\xspace}

\newcommand{\datastore}{\ST Scientific Datastore\xspace}
\newcommand{\computefarm}{The \ST Scientific Compute Farm\xspace}
\newcommand{\computefarmlong}{UCD CASL \ST Software and Data Compute Server
Farm\xspace}
\newcommand{\sensorfarm}{The \STSBs\xspace}

@@

  Sensor equipped devices typically communicate with a streaming packet-based 
  network protocol.
  The communication is focused on controlling the sensors and report the 
  sensors measurements.
  The \STSB is an example of such a device, it communicates with an arbitrary 
  computer system, the \STPU (running Linux or Windows, at the moment) over 
  a USB bus using a streaming packet-based protocol.
  The \STU is a custom-designed, but inexpensive, sensor platform: it is 
  composed by a one or more \STSBs paired with a general-purpose small-scale 
  compute node, the \STPU (including everything from PDAs to powerful 
  touchscreen-based portable computers).
  There are over a dozen sensors on the \STSB itself, and new sensors can be 
  added either using expansion slots on the \STSB, or adding them directly to 
  the USB bus.
  A third-party low-level library, provided by the USB chip vendor, the USB 
  chip that has been used in the \STSB, creates and parses
  generic USB communication packets.

  A custom medium-level Java driver creates and parses the \STSB streaming 
  communication packets. 
  This latter Java driver is formally specified, validated with rigorous
  unit testing, and verified using extended static checking. 
  Because of this process and combination of technologies, using this
  specification we are able to both communicate with a real \ST
  \textbf{and} generate a formally specified, executable, functional
  model-based specification for the protocol itself written in the
  Java Modeling Language.
  \todo executable? how? what does this mean?
  This protocol model is itself an executable
  stream simulator that can be used to develop new \ST user
  applications without having the \ST hardware at all. 
  Consequently,
  from this single specification, we can test that the \ST hardware
  communicates as expected, test that our \ST simulator communicates
  as expected, and simulate the \ST protocol stream itself.

@@

The \ST

@@

The communication between \STSB and \STPU is realized through an Universal 
Serial Bus (USB) interface.
On top of the USB serial channels a proprietary protocol is used, specified by 
FTDI\footnote{\myhref{http://www.ftdichip.com/}{FTDI web site}}; FTDI is the 
manufacturer of the USB controller chip used in the \STSB, the 
FT2232H\cite{ftdi_ft232h_2009}.

The proprietary protocol enables the control and management of some of the 
functionalities in FT232H chip (such as internal EEPROM read and write, reset 
the chip, etc.) as well as the definition of the communication protocol and 
properties to be used in the asynchronous channels (input and output channels) 
provided by FT232H.
These added functionalities are accessed via an Application Program Interface,
the API is provided by a driver developed directly by FTDI, which is available 
for various systems (Linux 32 and 64 bits, Mac OS, Windows)
\cite{ftdi_d2xx_api_2009}.

Finally, on top of the serial channels provided by the API, a custom protocol 
has been developed to manage the entire \STSB; that is, read sensor data 
observations, enable or disable sensors, configure onboard components 
as the Field-Programmable Gate Array (FPGA), and the micro processor; the 
protocol main function is to read sensor data.

The protocol is an asynchronous fixed length packed based protocol.
We will only examine the output sensor data packet, since it is the more used 
and interesting one.

The internal structure of the output sensor data packet reflects the board 
capabilities and sensors.
Basically a single packet has to accomodate various types of data:
\begin{itemize}
 \item Fat data rate streams: audio data, 48KHz or 96KHz.
 \item Medium data rate streams: 5KHz.
 \item Slow data rate streams: temperature, pression, accellerometer, etc.; all 
 these sensor have a data rate far lower than 5 KHz.
 \item Metadatas: metadatas on sensors and board state.
\end{itemize}

Even if we have to deal with data streams that have different data rates, the 
internal structure of the package (as we will see) is strictly fixed. 
This choice originates from a tradeoff between two constraints: the efficiency 
of the trasmission (redundant data to be minimized) and the complexity to build 
a package (complexity to be minimized).
The complexity minimization constraint is simple to understand: it is the FPGA 
on the \STSB that build the packet, but it has also to be used for various 
other specific tasks. 
Hence, it is important to maintain the building packet code on the FPGA as 
simple and short as possible, to leave enough space for the code implementing 
other tasks.
The Output sensor data packet structure is showed in Table 
\ref{tab:data_packet_structure}, the packet total length is 1024 bytes (512 16 
bit words).

\begin{table}[htbp]
\caption{Output sensor data packet structure.}
\label{tab:data_packet_structure}
\begin{center}
\begin{tabular}{|c|l|}\hline
\textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
0-1 & sentinel\\\hline
2-7 & slow data\\\hline
8-11 & metadata: board state\\\hline
12-13 & reserved\\\hline
14-15 & metadata: board state\\\hline
16 & reserved\\\hline
17-22 & subsequence 1\\\hline
\ldots & \ldots\\\hline
503-508 & subsequence 82\\\hline
\end{tabular}
\end{center}
\end{table}

The subsequence accomodate data from the fast and medium data rate streams 
(there are 4 fast data rate stream and 8 medium data rate stream channels). 
There are 82 subsequences in a packet.
The subsequence structure is shown in Table \ref{tab:subsequence_structure}.

\begin{table}[htbp]
\caption{Output sensor data packet subsequence structure.}
\label{tab:subsequence_structure}
\begin{center}
\begin{tabular}{|c|l|}\hline
\textbf{\textit{index}} & \textbf{\textit{function}}\\\hline
0 & metadata: subsequence description\\\hline
1 & medium data rate data sample\\\hline
2-5 & fast data rate data samples\\\hline
\end{tabular}
\end{center}
\end{table}

It is clear that, looking at the structure described and at the data rates 
frequencies, that there must be something subtle, to make all the figures to 
add up.
The subleties lies in two forms of ridundancies, for slow and mediumd data 
rates:

\begin{itemize}
  \item Slow data rates: slow data can be found at the beginning of each 
  packet, the data is simply repeated (that is, it is equal to the previous 
  packet in the stream) when no new data is available.
  \item Medium data rates: the 8 medium rate channels are multiplexed together 
  into the medium data rate data sample of each subsequence, which channel the 
  sample belongs to is a piece of info to be extracted from the subsequence 
  metadata. Besides, the medium data rate data sample could contain no data at 
  all, this info again is to be extracted from the metadata.
\end{itemize}
